Identification of Direct Cherenkov Pixels using Boosted Decision Trees

### 2016-06-28 report/draft1.tex
Backup 28/06/16

\documentclass{article}
\usepackage{graphicx}
\graphicspath{ {graphs/} }
\usepackage{geometry}
\geometry{a4paper, portrait, margin=1.0in}
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{amsfonts}
\begin{document}
\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

@@

@@

\section{HESS-type Event Reconstruction}

@@

For convenience, an offset $C=-6.5$ was omitted, yielding a simple exponential law. Counter-intuitively, it does in fact not matter whether the training values bear any close relation to the $True_DC$ values. The only values that we can use for reconstruction are our $candidate_{DC count}$, based on the DC candidate pixel and full-shower image. As long as these $candidate_{DC count}$ values are proportional to $Z^{2}$, the reconstruction method will remain valid. We train the regressor to interpolate the value of $True_{DC}$ using a number of variables, and most importantly, we include the True distance to the Core and the true energy among the regressor variables listed in Table \ref{tab:hess1regressor}.

@@

If suitably trained, the regressor would consider the core position as a factor, but also correlate it with other variable related to the image. Then, given an estimate of the core distance, the regressor is able to produce a value of $candidate_{DC count}$ with some small error, that is still proportional to $Z^{2}$. The training set of data consisted of one true DC pixel entry in each full shower image. To account for the differing hardware, a separate regressor was trained, as before, for the CT5 camera. The Feature Importances for the HESS1 and HESS2 regressor are listed in Table \ref{tab:hess1regressor} and Table \ref{tab:hess2regressor} respectively. There is again a discrepancy in variable importance between the two BDTs.

\begin{table}[h!]
  \centering
  \caption{Relative Feature Importance in HESS-1 Regressor BDT training}
  \label{tab:hess1regressor}
  \begin{tabular}{ccc}
    \toprule
    Variable & Relative Importance\\
    \midrule
    $Q_{DC}$ & 0.17\\
    $DC_{Count}$ & 0.16\\
    Image Amplitude & 0.15\\
    $Intensity_{N.N.min}$ & 0.14\\
    $Intensity_{N.N.max}$ & 0.14\\
    $Intensity$ & 0.13\\
    $Mean_{N.N}$ & 0.11\\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h!]
  \centering
  \caption{Relative Feature Importance in HESS-2 Regressor BDT training}
  \label{tab:hess2regressor}
  \begin{tabular}{ccc}
    \toprule
    Variable & Relative Importance\\
    \midrule
    Image Amplitude & 0.20\\
    $Intensity_{N.N.min}$ & 0.15\\
    $Q_{DC}$ & 0.15\\
    $Mean_{N.N}$ & 0.14\\
    $DC_{Count}$ & 0.13\\
    $Intensity_{N.N.max}$ & 0.13\\
    $Intensity$ & 0.13\\
    \bottomrule
  \end{tabular}
\end{table}

Repeating the comparison for the HESS1 regressor, we find for $\sigma_{rgrfit} = 0.35$, a significant improvement over $\sigma_{DCcount}$. Taking into account $\sigma_{STA}$, we find that ultimately $\sigma_{rgr1}=0.34$, a fairly large value that nonetheless should be acceptable in 4-telescope events. The comparative performance is recorded in Table \ref{tab:lpderror}. Instead of rejecting DC pixels which do not meet out $DC_{count}$ and $P_{signal}$ cuts, we can separately measure their deviation from the LPD. In this way, we can include the information from these DC pixels in later analysis, while accounting for the larger uncertainty in their LPD measurements. For images which do pass only the multiplicity cuts, but not the $P_{signal}$ or $DC_{count}$ cuts, we find that the regressor error is $\sigma_{rejected1}=0.32$.

@@

We can parameterise the error by considering the energy and distance to the core for a selection of DC pixels. We find that, for a fixed energy...

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{DCcounterrorhess1rgrbigtestdata}
\caption{The difference between true and reconstructed DC signal for events passing the cuts is shown in the top left. In the top right the same is shown for events which were rejected. The true DC signals for events which passed the cuts is plotted against the reconstructed signals in the lower left, and for rejected events in the lower right. As in \ref{fig:cutdistribution}, in all plots a green event is one in which the DC pixel has been correctly identified, while a red event is one that has been incorrectly identified.}
\label{fig:dcdiff}
\end{center}
\end{figure}

\subsection{Other Stuff Like}
Having obtained a parameterisation of the LPD, and and assoiated error in the case of a HESS-type array, the Charge reconstruction Technique can be applied to a simpified Monte Carlo simulation of the HESS array, developed as part of this analysis. Conducting a Monte Carlo simulation using the Gaussian-smeared LPD, we can measure the effectiveness of the LPD-reconstruction technique.

@@

As with energy, we wish to simulate a realistic spectrum for the first interaction heights. We must thus convert the uniform random number R to an interaction distance X. Using the standard atmospheric tables, the corresponding height value can be found using exponential interpolation.

@@

But you lose the tail..

As cited by the HESS study, a typical first interaction height is roughly 30km for a Cherenkov-Emitting event \cite{hess07}. With the assumed value of $x_{0}$, a simulation was conducted to determine the expected interaction heights for Cosmic Rays. It is found that the mean first interaction height of 36.9km and a median first interaction height of 35.2km for all cosmic rays. However, when Cherenkov-Emitting events are considered, we have both a median and mean first interaction height of 30.3km, which clearly agrees with our expectations.

@@

\subsection{First Interaction and Energy Saturation}
Using the theoretical LPD found from the Frank-Tamm formula, we see that this LPD will have a very abrupt discontinuity.

The mean first interaction height for all Cherenkov Emitting Cosmic Rays in the atmosphere is 40km, and neglecting variations in atmospheric density profiles around the Earth, this can be considered independent of experimental array. However, the multiplicity of event is determined by equipment efficiencies, altitude and telescope layout. When the HESS layout is simulated, we find that the 4 telescope events have a mean height of $h \approx 23 \pm 5$ km, as shown in \ref{fig:Hessheight}.

@@

Using basic Hillas reconstruction techniques we can find a rough estimate of the core position. For each telescope, we will obtain an expected direction to shower core, although it will be heavily Gaussian-smeared from the true direction. However, combining these values can restrict us to an expected target region. We consider a grid of spacing 5m in the xy plane, and with each telescope, consider the reconstructed direction to shower core. Allowing a certain angular deviation will give us a target region containing certain likely grid points. We progressively increase the angular deviation until we have found the twenty most likely core position points.

@@

\subsection{Log Likelihood Minimisation}
In order to fully reconstruct an event, we need to find the x/y core position, the Energy per Nucleon, the first interaction height and the charge. However, if one telescope in a five-telescope array does not observe DC light, this data point can be used to constrain the core position. Thus, for the LDF method to be applied, we require a minimum of five telescopes, four or more of which must image the DC light. We consider the amount of DC light that each telescope receives to be Gaussian, with an expectation of $\mu(X, Y, Z, height, Epn)$ and a standard deviation of $\sigma(\mu)$.

\[  P_{i} ( N_{i, Received} \mid X, Y, Z, height, Epn )  =  \frac{1}{\sqrt{2 \sigma^{2} \pi}} \exp(-\frac{(N_{i, Received} - \mu)^{2}}{2 \sigma^{2}}) \].

We then minimise the Log Likelihood function, assuming n is the total number of telescopes in the array. \[ - \ln(L) = - \sum_{i=1}^{n} \ln(P_{i}) =  \sum_{i=1}^{n} [ \frac{1}{2}\ln(2 \pi) + \ln(\sigma) + \frac{(N_{i, Received} - \mu)^{2}}{2 \sigma^{2}}]\]

@@

\subsection{Iterative Scanning Minimisation}
Through $\sigma_{Z}$ calculation, we find that the LDF method initially provides very poor charge reconstruction. This is as a result of varying Threshold Energies and the sharp drop in the LDF above the maximum radius, which means the Log Likelihood is frequently discontinuous within the parameter space. Consequently, a Minuit-type minimisation algorithm will only be able to find a local minimum near the starting values for the fit parameters.

To overcome this problem, we can iterate over a series of starting values for the parameters, with the aim of scanning the true minimum among the many minima found. To simplify matters, we can scan only the integer Z values over the range $ 20 \leq Z \leq 32 $, rather than considering the charge to be a free floating parameter.

In order to reduce the number of calculations needed, we can consider the core position region derived from camera images, as in Hillas Analysis. Each telescope will have a Gaussian smeared central axis indicating the direction of the core. We overlay the simulated telescope array with a grid of points with grid width 1m. Then we select all points satisfy the condition of lying within the shower axis direction confidence interval for every telescope. The angular width of the confidence interval from the shower axis will be determined by the precision of the telescopes.

Having determined the relevant Height and Energy ranges, we can then consider combinations of Energy and Height for which the Energy is above the Cherenkov Emission Threshold (and is saturated!). This gives us a second set of starting \textquoteleft Coordinates \textquoteright.

The Z value is fixed and the LL function is then minimised with the assigned starting values, with the other four variables allowed to float freely. Minimisation typically scans 13 Z values, 10 core position coordinates, and 50 Height/Energy coordinates, yielding $ 13 \times 10 \times 50 = 6500$ minimisations in total. Such a technique is resource intensive but reduces $\sigma_{Z}$ by a factor of 5 or more. (CHECK this quantitatively and get a graph yo. Want to see some plateauing of sigma Z!).

@@

\subsection{Saturation Region Energies}

@@

\section{Conclusion}
Preliminary results suggest that the LDF reconstruction technique will significantly improve charge reconstruction, to a level sufficient for cosmic ray abundance studies. However, reliance on high-multiplicity events means that although applicable to current experiments such as HESS, a new optimised telescope array would be required for a statistical analysis. Such an array may have a grid spacing of 20-50m, although further study is needed to determine the ideal layout.
\bibliographystyle{unsrt}
\bibliography{report}
\end{document}

>>>>

\documentclass{article}
\usepackage{graphicx}
\graphicspath{ {graphs/} }
\usepackage{geometry}
\geometry{a4paper, portrait, margin=1.0in}
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{pgfplotstable, booktabs}

\pgfplotstableset{
  	every head row/.style={before row=\toprule,after row=\midrule},
    every last row/.style={after row=\bottomrule}
	}

\begin{document}
\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

@@

We need four-telescope events or something.

@@

\section{HESS-type Event Simulation}

@@

For convenience, an offset $C=-6.5$ was omitted, yielding a simple exponential law. Counter-intuitively, it does in fact not matter whether the training values bear any close relation to the $True_DC$ values. The only values that we can use for reconstruction are our $candidate_{DC count}$, based on the DC candidate pixel and full-shower image. As long as these $candidate_{DC count}$ values are proportional to $Z^{2}$, the reconstruction method will remain valid. We train the regressor to interpolate the value of $True_{DC}$ using a number of variables, and most importantly, we include the True distance to the Core and the true energy among the regressor variables listed in Table \ref{tab:regressor}.

@@

If suitably trained, the regressor would consider the core position as a factor, but also correlate it with other variable related to the image. Then, given an estimate of the core distance, the regressor is able to produce a value of $candidate_{DC count}$ with some small error, that is still proportional to $Z^{2}$. The training set of data consisted of one true DC pixel entry in each full shower image.

The training sample contains, in equal proportion, elements Manganese (Z=25), Iron (Z=26) and Cobalt (Z=27). This to ensure that the regressor is well trained to distinguish between different cosmic ray charges, by providing variation of $Z^{2}$ for training. If the regressor is capable of distinguishing between these three elements, then this will serve as a sufficient proof of concept. We expect that, were an entire spectrum of cosmic ray elements simulated, a regressor could be trained with similar performance for a large charge range. The training of such a BDT was not done as part of this analysis, due to computing time restraints. However, in Monte Carlo simulations, the existence of a regressor will be assumed.

To account for the differing hardware, a separate regressor was trained, as before, for the CT5 camera. The Feature Importances for the HESS1 and HESS2 regressor are listed in Table \ref{tab:regressor}.. There is little discrepancy in variable importance between the two BDTs.

\begin{table}[h!]
  \centering
  \caption{Relative Feature Importance in  Regressor BDT training}
  \label{tab:regressor}
  \begin{tabular}{ccc}
    \toprule
    Variable & HESS-1 & HESS-2 \\
    \midrule
    Aspect Ratio & 0.12 & 0.12\\
    $Intensity_{N.N.min}$ & 0.11 & 0.12\\
    Image Amplitude & 0.11 & 0.11\\
    $Q_{DC}$ & 0.10 & 0.11\\
    $r_{core}$ & 0.10 & 0.10\\
    $Intensity_{N.N.max}$ & 0.10 & 0.10\\
    $Mean_{N.N}$ & 0.10 & 0.09\\
    Energy & 0.09 & 0.09\\
    $DC_{Count}$ & 0.09 & 0.09\\
    $Intensity$ & 0.08 & 0.08\\
    \bottomrule
  \end{tabular}
\end{table}

The features importances are, on the face of it, rather surprising. Ultimately, the true values for energy and $r_{core}$ are not the most important variables. We can conclude that the regressor has been well taught to indirectly infer the cosmic ray charge, and account for this by correlating values to the true core distance/energy. The Aspect Ratio is heavily dependent on distance to core, but also on the charge. Thus, in combination with $r_{core}$, the charge can be reasonably inferred. Similarly the Image Amplitude is relied upon to infer energy, and in conjunction with the estimated energy, can indicate charge. It is encouraging to see that the regressor relies more heavily on Image Amplitude and Aspect Ratio, because these two values are known with a much smaller uncertainty than the core position/energy.

Repeating the comparison for the HESS1 regressor, we find for $\sigma_{rgr} = 0.14$, a significant improvement over $\sigma_{DCcount}$. The plotted LPD is seen in Figure \ref{fig:corsikalpd1}. Through use of a well-defined law for the training values, we have eliminated the influence of $\sigma_{STA}$ on the training process. The comparative performance is recorded in Table \ref{tab:lpderror}.

Instead of rejecting DC pixels which do not meet out $DC_{count}$ and $P_{signal}$ cuts, we can separately measure their deviation from the LPD. In this way, we can include the information from these DC pixels in later analysis, while accounting for the larger uncertainty in their LPD measurements. For images which do pass only the multiplicity cuts, but not the $P_{signal}$ or $DC_{count}$ cuts, we find that the regressor error is $\sigma_{rejected1}=0.15$. We can thus make use of these points, because they follow an almost identical distribution.

STUFF

@@

\section{Monte Carlo Simulation}
Having obtained a parameterisation for both the full and DC LPD, as well as the associated error in the case of a HESS-type array, the Charge reconstruction Technique can be applied to a simpified Monte Carlo simulation of the HESS array, developed as part of this analysis. Conducting a Monte Carlo simulation using the Gaussian-smeared LPD, we can measure the effectiveness of the LPD-reconstruction technique. The simulated Cosmic Rays must be high-multiplicity, and have a realistic distribution of energy and interaction height values.

@@

As with energy, we wish to simulate a realistic spectrum for the first interaction heights. We must thus convert the uniform random number R to an interaction distance X. Using the standard atmospheric tables, the corresponding height value can be found using exponential interpolation. The standard atmospheric data is included in the appendix.

@@

Interpolating the value of the height neglects the fact that some cosmic rays which would not interact with the atmosphere before reaching the ground. The integrated interaction lengths are equal to about 850 at an altitude comparable to the HESS site. There is thus a probability of reaching the ground equal to $P = exp(\frac{-850}{12}) \approx 1.73 \times 10 ^{-31}$. It is thus reasonable to ignore this tiny fraction of events, and instead assume that every cosmic ray will interact at some point in the atmosphere.

As cited by the HESS study, a typical first interaction height is roughly 30km for a Cherenkov-Emitting event \cite{hess07}. With the assumed value of $x_{0}$, a simulation was conducted to determine the expected interaction heights for Cosmic Rays. It is found that the mean first interaction height of 36.9km and a median first interaction height of 35.2km for all cosmic rays. However, when Cherenkov-Emitting events are considered, we have both a median and mean first interaction height of 30.3km, which clearly agrees with our expectations. Considering solely high-multiplicity events, we find that these have a mean height of $h \approx 23 \pm 5$ km, as shown in Figure \ref{fig:Hessheight}.

@@

MOAR!!!

@@

Using basic Hillas reconstruction techniques we can find a rough estimate of the core position. For each telescope, we will obtain an expected direction to shower core, although it will be heavily Gaussian-smeared from the true direction. However, combining these values can restrict us to an expected target region. We consider a grid of spacing 5m in the xy plane, and with each telescope, consider the reconstructed direction to shower core. Allowing a certain angular deviation will give us a target region containing certain likely grid points. We progressively increase the angular deviation until we have found the twenty most likely core position points. It is important to note that these core positions are simply starting points for a minimisation. The minimisation itself is allowed to float freely anywhere in the simulated target grid of $300 \times 300 = 90000 m^{2}$.

@@

\subsection{Log Likelihood Minimisation}
Having estimated the rough core position and energy, we would at this point use our regressor to determine the values of $candidate_{DCcount}$, based on the BDT-identified DC pixel and image parameters. Having done so, we know that these values would follow the regressor LPD. For the Monte-Carlo simulation, this step is bypassed, and the regressor LPD is instead directly simulated with a standard Gaussian smearing. Having quantified the expected deviation from the LPD under these conditions, this is a reasonable simplification that massively reduces the computing requirements for event simulation. This in turn enables us to consider a greatly expanded dataset for reconstruction attempts.

In order to fully reconstruct the events, we need to find the x/y core position, the Energy per Nucleon, the first interaction height and the charge. We again consider the amount of DC light that each telescope receives to be Gaussian, with an expectation of $\mu(X, Y, Z, height, Epn)$ and a standard deviation of $\sigma(\mu)$. Ultimately the energy dependence of the LPD is almost negligible over the energy range, because variable $r_{max}$ is only weakly dependent on the energy. It thus makes little sense to attempt to reconstruct the energy a second time, when we would expect no discernible improvement over the reconstruction solely based on the Full Shower LPD that was performed before. We thus fix the cosmic ray energy as that found before, and instead consider a four-dimensional minimisation in Charge, x and y core position, and first interaction height.

We then minimise the Log Likelihood function, where the total log likelihood for each telescope is equal to a sum of the full-shower LPD log likelihood and the DC LPD log likelihood. We are thus minimising a Likelihood function with nine or ten measurements from two broadly independent distributions. We consequently expect an improvement in the position reconstruction, as well as a measurement of the charge and the first interaction height.

As before, we find that the discontinuous LPD leads to a frequent identification of local rather than global minima. To overcome this problem, we can iterate over a series of starting values for the parameters, with the aim of scanning the true minimum among the many minima found. In addition to the grid of likely positions identified for the Full-Shower LPD minimisation, we can can scan the integer Z values over the range $ 20 \leq Z \leq 32 $. It should be emphasised that the charge Z is treated as a free-floating parameter during the minimisation, and is able to vary freely within the range 16-36. Similarly, three first interaction heights in the range 20-30 km are scanned, but the value is able to vary freely in the range 15-65 km. It is purely due to computer resource restraints that \textquoteleft likely coordinates' in the parameter space are scanned as a starting point for a minimisation, rather than a uniformly distributed scan. Despite the simplifications, each minimisation process will still undergo $12 \times 20 \times 3 = 720$ iterations before a final reconstruction is complete.

\subsection{Variable Resolution}
If we compare the deviation from true core distance to each telescope, we can quantify the improvcement in core position reconstruction from minimisation based solely on the full-shower LPD. This is shown in Figure \ref{fig:coredistance}, and is directly comparable with the preliminary results in Figure \ref{fig:rawcoredistance}. As expected, the core distance resolution is reduced to X and Ym, and clear improvement over the previous measurements of core position.

As an alternative metric for assessing reconstruction accuracy, we can calculate the absolute distance from the reconstructed core to the true core. The results of a binning of this distribution are shown in Figure \ref{fig:position}. It is found that, for both four and five telescope events, the core reconstruction is approximately $\sigma_{core} \approx 10m$. Most core positions are reconstructed less than 10m from the true position. This is extremely good, especially in comparison to the Hillas Reconstruction, where a core resolution of Y is more typical AND CITE!.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{coredistance}
\caption{The distance from true core to the core reconstructed with the full-LPD is shown above. For four-telescope events, the standard deviation from true position is Xm. }
\label{fig:coredistance}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{position}
\caption{The distance from true core position to the reconstructed core position is shown above. For four-telescope events, the standard deviation from true position is Xm.}
\label{fig:position}
\end{center}
\end{figure}

@@

@@

\subsection{Saturation Region Energies}
Hillas reconstruction better
Combine with Hillas reconstruction

@@

\section{Conclusion}
Preliminary results suggest that the LDF reconstruction technique will significantly improve charge reconstruction, to a level sufficient for cosmic ray abundance studies. However, reliance on high-multiplicity events means that although applicable to current experiments such as HESS, a new optimised telescope array would be required for a statistical analysis. Such an array may have a grid spacing of 20-50m, although further study is needed to determine the ideal layout.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{DCcounterrorhess1rgrbigtestdata}
\caption{The difference between true and reconstructed DC signal for events passing the cuts is shown in the top left. In the top right the same is shown for events which were rejected. The true DC signals for events which passed the cuts is plotted against the reconstructed signals in the lower left, and for rejected events in the lower right. As in \ref{fig:cutdistribution}, in all plots a green event is one in which the DC pixel has been correctly identified, while a red event is one that has been incorrectly identified.}
\label{fig:dcdiff}
\end{center}
\end{figure}
\section{Bibliography}
\bibliographystyle{unsrt}
\bibliography{report}
\section{Appendix}
\subsection{Atmospheric Data Table for CORSIKA (Windhoek, Namibia)}
\pgfplotstabletypeset{data/atmprof10.dat}
\end{document}
